\documentclass[msc,cs,parskip,leftchapter,logo,twoside,abbrevs,11pt]{infthesis}

\usepackage[square,authoryear]{natbib}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{eulervm}
\usepackage{color}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{lipsum}
\usepackage{verbatimbox}
\usepackage{gnuplottex}
\usepackage{amssymb}
\usepackage{wrapfig}
\usepackage{tikz}
\usepackage[toc,page]{appendix}
\usetikzlibrary{arrows,automata}

% set colour for listings
\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{mymauve}{rgb}{0.58,0,0.82}

\lstset{ %
  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\ttfamily,        % the size of the fonts that are used for the code
  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
  breaklines=true,                 % sets automatic line breaking
  captionpos=b,                    % sets the caption-position to bottom
  commentstyle=\color{mygreen},    % comment style
  deletekeywords={...},            % if you want to delete keywords from the given language
  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
  frame=single,                    % adds a frame around the code
  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  keywordstyle=\color{blue},       % keyword style
  language=Java,                 % the language of the code
  morekeywords={def,function,assert,aspect,pointcut,before,transparent,co},
  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=5pt,                   % how far the line-numbers are from the code
  numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
  showstringspaces=false,          % underline spaces within strings only
  showtabs=false,                  % show tabs within strings adding particular underscores
  stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
  stringstyle=\color{mymauve},     % string literal style
  tabsize=4,                       % sets default tabsize to 2 spaces
  title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}

% include subsubsections in toc
\setcounter{secnumdepth}{5}
\setcounter{tocdepth}{5}

% Change font to palatino
% Euler for math | Palatino for rm | Helvetica for ss | Courier for tt
\renewcommand{\rmdefault}{pplx} % rm
\linespread{1.05}        % Palatino needs more leading

\title{Parallelism Detection using Dynamic\\Instrumentation in a Virtual Machine}
\author{Christopher Edward Atkin-Granville}

\abstract{Contemporary computers are increasingly based around highly parallel architectures through chip multiprocessors, instruction-level parallelism and graphics processing units with potentially thousands of cores. Despite this, many popular programs are based around a sequential programming paradigm.

This project investigates the use of the Graal compiler infrastructure in order to dynamically profile running programs within the Java Virtual Machine and determine which hot loops are good candidates for automatic parallelism transformations, possibly JIT recompilation to an OpenCL target. We show that, currently, Graal is not yet mature enough to support the transformations required. However, in the future, it will becomes mature enough and our framework is a simple `drop-in' to the Graal compilation procedure to add instrumentation to programs.

We consider two main approaches to trace collection: exact approaches and probabilistic approaches based on bloom filters. A new benchmark has been created that allows for fine-tuning of the number and types of dependencies, as well as the amount of computation associated with each operation (in the form of a number of floating-point operations) which allows us to analyse the efficacy of each approach without the disadvantages of real-world benchmarks (\ie, where we cannot know ahead-of-time how many dependencies there are, and so cannot evaluate the performance of each approach).

The findings presented by this dissertation show that not only is it feasible to use bloom filters for these kinds of analysis, but it also advantageous to do so over hash sets. With the correct parameters, bloom filters show a small memory use increase over uninstrumented programs, and a relatively small increase in the execution time of between a factor of 1.4 and 2 times slower.}

\begin{document}
	\begin{preliminary}
		\maketitle
		\begin{acknowledgements}
		I wish to thank my supervisors, Dr. Christophe Dubach and Dr. Bj\"{o}rn Franke for their insightful and valuable contributions to the project.

		Also in need of thanks are my friends and especially my parents, Sandra and Ian who have supported me throughout my entire University career.
		
		Lastly, I would like to thank all those folks on the Graal mailing list who answered all my questions regarding the system.
		\end{acknowledgements}
		\standarddeclaration
		\dedication{To my grandfathers, Leslie and John.}
		\tableofcontents
		\listoffigures
		%\listofalgorithms
	\end{preliminary}

	\include{chapters/introduction}
	%\include{chapters/background}
	\include{chapters/related-work}
	\include{chapters/graal}
	\include{chapters/instrumentation}
	\include{chapters/runtime-library}
	\include{chapters/methodology}
	%\include{chapters/parametric-benchmarks}
	\include{chapters/results}
	%\include{chapters/optimisations}
	\include{chapters/conclusion}
	
	\appendix
	\include{chapters/appendix-precision}
	\include{chapters/appendix-time}
	\include{chapters/appendix-memory}	
	\include{chapters/appendix-disable}

	\singlespace
	\bibliographystyle{unsrtnat}
	\bibliography{library}
\end{document}